---
title: "Pstat 174 Final Project"
author: "Skyler Ataide"
date: "6/2/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tsdl)
tsdl

tsdl <- subset(tsdl,"Health")
tsdl

attributes(tsdl[[6]])

View(tsdl[[6]])

num_cases <-as.vector(tsdl[[6]]) 

plot.ts(num_cases)

```

library(tsdl)
tsdl

tsdl <- subset(tsdl,"Health")
tsdl

attributes(tsdl[[6]])

View(tsdl[[6]])

num_cases <-ts(as.vector(tsdl[[6]]))

plot.ts(num_cases, ylab = "Number of Chickenpox Cases")

nt = length(num_cases)

## Add trend line to data plot:
fit <- lm(num_cases ~ as.numeric(1:nt)); abline(fit, col="red", ylab = "Number of Chickenpox Cases")

mean(num_cases)

## Add mean (constant) to the data plot:
abline(h=mean(num_cases), col="blue")

## Plot chickenpox data with years on x=axis: 
tsdata <- ts(num_cases, start = c(1931,1), end = c(1972,2), frequency = 12)

ts.plot(tsdata, main = "Raw Data Chickenpox Cases NYC 1931-1972", ylab="Number of Chickenpox Cases")

fit <- lm(tsdata ~ as.numeric(1:length(tsdata))); abline(fit, col="red")

abline(h=mean(tsdata), col="blue")




## Partition dataset to two parts for model training and model validation; work with training set:
nc.train = ts(num_cases[c(1:444)]) # Training Set
nc.test = ts(num_cases[c(445:498)]) # Test Set 


## Chickenpox Training (truncated) Set: 444 observations
plot.ts(nc.train,  main = "Training Set:Truncated Chickenpox Cases NYC", ylab="Number of Chickenpox Cases")


fit <- lm(nc.train ~ as.numeric(1:length(nc.train))); abline(fit, col="red")

abline(h=mean(num_cases), col="blue")

## Immediate Observations: Non-constant of variance and mean. No obvious increasing or decreasing trend. Highly Non-Stationary. 

## Confirm non-stationarity of original data (truncated)(nc.train)
## Plot Histogram Training Set:
hist(nc.train, col="dark blue", xlab="", main="Histogram; Truncated Chickenpox Data")
## Observations: Histogram is badly skewed. 

## Plot acf of Training Set
acf(nc.train,lag.max=NULL, main="ACF of the Truncated ChickenPox Data")
## Observations: Seasonality is obvious, at lag 12 it starts to repeat the previous cycle. ACFs remain large. 

pacf(nc.train,lag.max=NULL, main="PACF of the Truncated ChickenPox Data")


## Decomposition of the Log-Transformed truncated data: 
library(ggplot2)
library(ggfortify)

y <- ts(as.ts(nc.train.stat), frequency = 12)
decomp <- decompose(y, type = "additive")
plot(decomp)

## Decomposition Observations: the dataset has strong seasonality pattern. the lower bound of the data is constant, whereas the peak is changing.  there is a overall decreasing trend in the pattern. The data itself is not stationary as the variance is changing.

## Box-Cox Transformation:
bcTransform <- boxCox(nc.train~ as.numeric(1:length(nc.train)), main="Box-Cox Transformation")

## Command to give the value of lambda in the Box-Cox plot. 
bcTransform$x[which(bcTransform$y == max(bcTransform$y))] 
## lambda = 0.3434343

## Perform transformations, plot transformed data, histograms:
lambda=bcTransform$x[which(bcTransform$y == max(bcTransform$y))]
nc.train.bc = ts((1/lambda)*(nc.train^lambda-1))
nc.train.log <- ts(log(nc.train))

## Plot of truncated data after Box-Cox Transformation
plot.ts(nc.train.bc, main="Box-Cox Transformation: Truncated Chickenpox Data", ylab="Number of Chickenpox Cases")
fit <- lm(nc.train.bc ~ as.numeric(1:length(nc.train.bc))); abline(fit, col="red") 
mean(nc.train.bc)
abline(h=mean(nc.train.bc), col="blue")

## Histogram of truncated data after Box-Cox Transformation
hist(nc.train.bc, col="blue", xlab="", main="Histogram; Box-Cox Transformation")
var(nc.train.bc)

## Variance = 72.63398

## Plot of truncated data after Log Transformation
plot.ts(nc.train.log, main="Log Transformed Truncated Chickenpox Data", ylab="Number of Chickenpox Cases")

fit <- lm(nc.train.log ~ as.numeric(1:length(nc.train.log))); abline(fit, col="red") 
mean(nc.train.log)
abline(h=mean(nc.train.log), col="blue")

var(nc.train.log)
## Observations: Variance = 1.099665

## Histogram of the truncated data after Log Transform: 
hist(nc.train.log, col="blue", xlab="", main="Histogram; Log Transformation")


## Differencing of the Log-Transformed truncated data: 
## Differencing at Lag 12:
nc.train.log12 <- ts(diff(nc.train.log, lag=12))
plot.ts(nc.train.log12, main="Log Transformed Data Differenced at Lag 12")

fit <- lm(nc.train.log12 ~ as.numeric(1:length(nc.train.log12))); abline(fit, col="red") 
mean(nc.train.log12)
abline(h=mean(nc.train.log12), col="blue")

var(nc.train.log12)
## Variance = 0.1759289 log transformed and differenced at lag 12

## Differencing apt.log at Lag 12 and Lag 1: 
nc.train.stat <- ts(diff(nc.train.log12, lag=1))
plot.ts(nc.train.stat, main="Log Transformed Data Differenced at Lags 12 & 1")

fit <- lm(nc.train.stat ~ as.numeric(1:length(nc.train.stat))); abline(fit, col="red")

mean(nc.train.stat)
abline(h=mean(nc.train.stat), col="blue")

var(nc.train.stat)
## Observations: Variance = 0.09697977 LOWEST VARIANCE YET! No more seasonality. No more Trend. The data looks stationary, but now we must check the acf's to confirm. 


## Plot of ACF of log-transformed truncated data: 
acf(nc.train.log, lag.max=NULL, main="ACF of the log(U_t)")
## Observations: Non-stationarity. Seasonality is apparent. 

## Plot of ACF of log-transformed truncated data Differenced at lag 12: 
acf(nc.train.log12, lag.max=NULL, main="ACF of the log(U_t), differenced at lag 12")
## Observations: Seasonality, non-stationarity. 

## Plot of ACF of log-transformed truncated data Differenced at lag 12 and at Lag 1:
acf(nc.train.stat, lag.max=NULL, main="ACF of the Log Transformed Data, Differenced at Lags 12 & 1")
## Observations: ACF decay corresponds to a stationary process

## Decision: Work with the log-transformed truncated data differenced at lag 12 and lag 1. 

par(mfrow=c(1,1))

## Compare histograms of log-transformed truncated data:
## Histogram of log-transformed truncated data:
hist(nc.train.log, col="blue", xlab="", main="Histogram; Log Transformation")

##Histogram of log transformed truncated data differenced at lag 12
hist(nc.train.log12, col="blue", xlab="", main="Histogram; Log Transform, Differenced Lag 12")

## Histogram of log-transformed truncated data differenced at lag 12 and lag 1:
hist(nc.train.stat, col="blue", xlab="", main="Histogram; Log Transform Differenced Lags 12 & 1")
## Observations: This histogram looks symmetric and almost Gaussian. 

## Histogram of transformed and differenced data with normal curve:
hist(nc.train.stat, density=20,breaks=20, col="blue", xlab="", prob=TRUE, main="Log Transform Differenced Lags 12 & 1 + Normal Curve")
m<-mean(nc.train.stat)
std<- sqrt(var(nc.train.stat))
curve( dnorm(x,m,std), add=TRUE )

## Plot PACF:
pacf(nc.train.stat, lag.max=NULL, main="PACF of the Log Transformed Data, Differenced at Lags 12 & 1")

auto.arima(nc.train.log, D=1, d=1)


test(nc.train.stat)

arima(nc.train.log, order=c(0,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -15.53

arima(nc.train.log, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -22.26

arima(nc.train.log, order=c(0,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -55.02

arima(nc.train.log, order=c(0,1,4), seasonal = list(order = c(0,1,2), period = 12), method="ML")
## aic = -55.48

arima(nc.train.log, order=c(1,1,0), seasonal = list(order = c(0,1,0), period = 12), method="ML")
## aic = 216.47

arima(nc.train.log, order=c(2,1,0), seasonal = list(order = c(0,1,0), period = 12), method="ML")
## aic = 211.87

arima(nc.train.log, order=c(3,1,0), seasonal = list(order = c(0,1,0), period = 12), method="ML")
## aic = 209.21

arima(nc.train.log, order=c(1,1,0), seasonal = list(order = c(1,1,0), period = 12), method="ML")
## aic = 90.5

arima(nc.train.log, order=c(2,1,0), seasonal = list(order = c(1,1,0), period = 12), method="ML")
## aic = 80.84

arima(nc.train.log, order=c(3,1,0), seasonal = list(order = c(1,1,0), period = 12), method="ML")
## aic = 81.95

arima(nc.train.log, order=c(2,1,0), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -7.85

arima(nc.train.log, order=c(3,1,0), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -7.75

arima(nc.train.log, order=c(0,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -15.53

arima(nc.train.log, order=c(1,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -55.9

arima(nc.train.log, order=c(2,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -57.19

arima(nc.train.log, order=c(3,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -55.27

arima(nc.train.log, order=c(1,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -53.92

arima(nc.train.log, order=c(2,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -55.25

arima(nc.train.log, order=c(3,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -55.16

arima(nc.train.log, order=c(0,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -55.02

arima(nc.train.log, order=c(1,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## aic = -54.68

arima(nc.train.log, order=c(2,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML")

AICc(arima(nc.train.log, order=c(2,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML"))
## aic = -55.6


arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## AIC = -62.85

arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(0,1,1), period = 12), method="ML")
## AIC = -57.32

arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(1,1,1), period = 12), method="ML")

AICc(arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(1,1,1), period = 12), method="ML"))

## AIC = -65.36

arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(1,1,1), period = 12), method="ML")

AICc(arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(1,1,1), period = 12), method="ML"))
## AIC = -65.11


arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(2,1,2), period = 12), method="ML")
## AIC = -67.09

#### MODEL B
arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(2,1,2), period = 12), fixed = c(NA,NA,NA,NA,NA,0,NA,NA,NA,0,NA,NA), method="ML")

AICc(arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(2,1,2), period = 12), fixed = c(NA,NA,NA,NA,NA,0,NA,NA,NA,0,NA,NA), method="ML"))
## aic = -69.08

arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(2,1,2), period = 12), method="ML")

AICc(arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(2,1,2), period = 12), method="ML"))
## aic = -67.7

#### MODEL A
arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(2,1,2), period = 12), fixed = c(0, NA, NA, NA,NA,NA,NA,NA,0,NA,NA), method="ML")

AICc(arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(2,1,2), period = 12), fixed = c(0, NA, NA, NA,NA,NA,NA,NA,0,NA,NA), method="ML"))
## aic = -70.26

arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(2,1,1), period = 12), method="ML")
## AIC = -57.32





fit <- arima(nc.train.log, order=c(4,1,4), seasonal = list(order = c(2,1,2), period = 12), fixed = c(NA,NA,NA,NA,NA,0,NA,NA,NA,0,NA,NA), method="ML")

res <- residuals(fit)

hist(res,density=20,breaks=20, col="blue", xlab="", prob=TRUE, main = "Histogram of Residuals Model B")

m <- mean(res)
std <- sqrt(var(res))
curve( dnorm(x,m,std), add=TRUE )

m

plot.ts(res,  main = "Time Series Plot Residuals Model B", ylab="Residuals")

fitt <- lm(res ~ as.numeric(1:length(res))); abline(fitt, col="red")

abline(h=mean(res), col="blue")

qqnorm(res,main= "Normal Q-Q Plot for Model B")

qqline(res,col="blue")

acf(res, lag.max=20, main = "ACF Residuals Model B")

pacf(res, lag.max=20, main = "PACF Residuals Model B")

shapiro.test(res)

Box.test(res, lag = 21, type = c("Box-Pierce"), fitdf = 10)

Box.test(res, lag = 21, type = c("Ljung-Box"), fitdf = 10)

Box.test(res^2, lag = 21, type = c("Ljung-Box"), fitdf = 0)

acf(res^2, lag.max=NULL)

ar(res, aic = TRUE, order.max = NULL, method = c("yule-walker"))

install.packages("forecast")
library(forecast)



fit.A <- arima(nc.train.log, order=c(3,1,4), seasonal = list(order = c(2,1,2), period = 12), fixed = c(0, NA, NA, NA,NA,NA,NA,NA,0,NA,NA), method="ML")

forecast(fit.A) 

pred.tr <- predict(fit.A, n.ahead =12)

U.tr= pred.tr$pred + 2*pred.tr$se 

L.tr= pred.tr$pred - 2*pred.tr$se

ts.plot(nc.train.log, xlim=c(300,length(nc.train.log)+12), ylim = c(min(nc.train.log),max(U.tr)), main="Zoomed Forecast of Transformed Data Using Model A")

lines(U.tr, col="blue", lty="dashed")

lines(L.tr, col="blue", lty="dashed")

points((length(nc.train.log)+1):(length(nc.train.log)+12), pred.tr$pred, col="red", pch=18)



pred.orig <- exp(pred.tr$pred)

U= exp(U.tr)

L= exp(L.tr)

ts.plot(nc.train, xlim=c(1,length(nc.train)+12), ylim = c(min(nc.train),max(U)))

lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")

points((length(nc.train)+1):(length(nc.train)+12), pred.orig, col="red",pch=18)



ts.plot(nc.train, xlim = c(400,length(nc.train)+12), ylim = c(1,max(U)), main="Zoomed Forecast of Original Data Using Model A")

lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")

points((length(nc.train)+1):(length(nc.train)+12), pred.orig, col="red", pch=18)



ts.plot(num_cases, xlim = c(400,length(nc.train)+12), ylim = c(1,max(U)), col="dark green", ylab = "Number of Cases of Chickenpox", main="Zoomed Forecast with True Values from Test Set")


lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")

points((length(nc.train)+1):(length(nc.train)+12), pred.orig, col="red", pch=18)








ts.plot(num_cases, xlim = c(300,length(nc.train)+54), ylim = c(300,max(U)),col="red")





ts.plot(num_cases, xlim = c(200,length(nc.train)+12), ylim = c(250,max(U)), col="red")

lines(U, col="blue", lty="dashed")
lines(L, col="blue", lty="dashed")

points((length(nc.train)+1):(length(nc.train)+12), pred.orig, col="green")
points((length(nc.train)+1):(length(nc.train)+12), pred.orig, col="black")




pacf(nc.train.log, lag.max=NULL, main="PACF of the ln(U_t)")

adf.test(nc.train.stat)

auto.arima(nc.train.stat)

auto.arima(nc.train.log)





## Exploring the ACF and PACF gives a list of candidate models to try: 
arima(apt.log, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML")
AICc(arima(apt.log, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), method="ML"))

arima(apt.log, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), fixed = c(NA, 0, NA, NA), method="ML")
AICc(arima(apt.log, order=c(0,1,3), seasonal = list(order = c(0,1,1), period = 12), fixed = c(NA, 0, NA, NA), method="ML"))


















arima(nc.train.stat, order=c(0,0,1), seasonal = list(order = c(0,0,0), period = 12), method="ML")

AICc(arima(nc.train.stat, order=c(0,0,1), seasonal = list(order = c(0,0,0), period = 12), method="ML"))
## AIC = 217.098


arima(nc.train.log, order=c(3,1,2), seasonal = list(order = c(0,1,1), period = 12), method="ML")
##AIC = -56!! 

AICc(arima(nc.train.stat, order=c(3,0,2), seasonal = list(order = c(0,0,1), period = 12), method="ML"))
## AIC = -55.65978

arima(nc.train.stat, order=c(3,0,1), seasonal = list(order = c(0,0,1), period = 12), method="ML")


arima(nc.train.log, order=c(3,0,4), seasonal = list(order = c(0,0,1), period = 12), method="ML")
## AIC -63.61

arima(nc.train.stat, order=c(3,0,2), seasonal = list(order = c(1,1,1), period = 12), method="ML")

arima(nc.train.stat, order=c(4,0,1), seasonal = list(order = c(0,0,1), period = 12), method="ML")

AICc(arima(nc.train.stat, order=c(4,0,1), seasonal = list(order = c(0,0,1), period = 12), method="ML"))
## AIC = -54.78366






